Using best practices for forecasting biodiversity changes in breeding birds

### 2017-07-11 manuscript/draft.Rmd
evaluation & results

**Species Distribution Models.** Species distribution models (SDMs) predict
individual species' occurrence probabilities using environmental variables.
Species-level models are used to predict richness by summing the predicted
probability of occupancy across all species at a site; following [[Calabrese et
al.]], we calculated the uncertainty in our richness estimate by treating
richness as a sum over independent Bernoulli random variables (one per species).
This avoids known problems with the use of thresholds for determining whether or
not a species will be present at a site [[refs]]. We represented SDMs in this
analysis using random forests, a common approach in the species distribution
modeling literature based on fitting hundreds of independent regression trees to
randomly-perturbed versions of the data [[refs; would be good to provide a
source that says these are a good choice]]. Each of the 500 trees in our
species-level random forests used a different Monte Carlo estimate of the
observer effects as a predictor variable.

@@

We evaluated the performance of each model for forecasting using the data for
each year between `r settings$last_train_year` and `r settings$end_yr`. We used
three metrics for evaluating performance: 1) root-mean-square error (RMSE) to
determine how far, on average, the models' predictions were from the observed
value; 2) the 95% prediction interval coverage to determine how well the models
did at predicting uncertainty; and 3) deviance as an integrative measure of fit
incorporating good point estimates, precision, and coverage. When a model was
run multiple times for the purpose of correcting for observer effects, we used
the mean of those runs' point estimates as our final point estimate and we
calculated the uncertainty using the law of total variance (i.e.  the average of
the model runs' variance, plus the variance in the point estimates). In addition
to looking at forecast performance in general we evaluated how performance
changed as the time horizon of forecasting increased by plotting performance
metrics against the forecast horizon.

Variance in species richness within sites was lower than predicted by several
common count models (i.e. richness was underdispersed), so these count models
would have difficulty fitting the data. Since richness also had a relatively
large mean ([[N]]) and all models already produce continuous richness estimates
(due to summing probabilities in the (J)SDM case), we defined model performance
for all models in terms of continuous Gaussian errors instead of using discrete
count distributions.

@@

When comparing forecasts for richness across sites all methods performed well
(Figure 2; R^2 XX-XX%). However stacked and joint SDMs and the macroecological
model both failed to successfully forecast the highest richness sites resulting
in a noteable clustering of predicted values near ~60 species and the poorest
performance in the cross-site comparison (R^2 XX-XX%).

For most combinations of metric and forecast time lag, the best-performing model
for forecasting within site variability was either the "average" baseline model
or the ARIMA model (Figure 3). In general, each model's error increased
more-or-less linearly with time, although the slopes and initial error rates
varied across models (Figure 3). The site-level models (i.e. the baselines and
the ARIMA methods) tended to do very well (Figure 3) on short time horizons,
especially in terms of RMSE.

In general, the models produced confidence intervals that were too narrow,
indicating overconfident predictions. Of these, the random forest-based SDM
stack was the most overconfident. As noted elsewhere [@harris2015], these models
are generally far too confident, with observed richness values frequently
falling well outside their confidence limits (Figure 3C). The "naive" model was
the only model whose confidence intervals were too wide (Figure 3), which can be
attributed to the rapid rate at which these intervals expand as the time lag
increases (Figure 1). The JSDM and macroecological model has the best coverage.

RMSE and deviance produced qualitatively similar results in most cases, with the
notable exception of the stacked SDMs (Figure 3A and 3B). Deviance, unlike RMSE,
depends on the full predictive distribution of a model, including the model's
estimated uncertainty. As a result, the stacked SDMs' deviance was notably
higher than the next-worst model even though its RMSE was not unusual. On the
other hand, the richness-GBM model and especially the mistnet JSDM had
reasonably well-calibrated uncertainty estimates, even in the presence of
uncorrected observer biases (Figure 3C). This prevented their deviance from
rising as quickly as the site-level models when the observer population changed
and resulted in better comparative performance for these models using deviance
(Figure 3B).

Accounting for the uncertainty associated with differences among observers
resulted in improvements to all metrics of fit (Figure 4). The one exception was
RMSE of the "Naive" model, which remained unchanged. The improvements in point
estimates primarily resulted from large improvements in a small number of
forecasts, presumably ones where a large difference between observers resulted
in a large shift in predicted richness values that the models could not
otherwise accommodate. Most forecasts exhibited only small changes in
RMSE. However these changes were sufficient to reorder which models performed
best overall. For example, the "naive" model generally had the lowest average
error when observer effects were not explicitly accounted for, switching to the
average and ARIMA models once observer differences were corrected. Improvements
in coverage were notable for all models except the naive and JSDM, meaning that
incorporating observer effects resulted in improved quantification of
uncertainty. The lack of change in the mistnet JSDM seems to indicate that its
latent variables were already accounting for observer-related uncertainty. In
combination changes in point estimates and coverage resulted in changes in the
best models based on deviance at different time scales. The macroecological
model and the JSDM appeared to narrowly outperform the average and ARIMA models
when the forecast horizon was greater than five years when not correcting for
observers, but after the correction the average and ARIMA models maintained a
small average advantage out to at least ten years.

>>>>

**Species Distribution Model.** Species distribution models (SDMs) predict
individual species' occurrence probabilities using environmental variables.
Species-level models are used to predict richness by summing the predicted
probability of occupancy across all species at a site; following [[Calabrese et
al.]], we calculated the uncertainty in our richness estimate by treating
richness as a sum over independent Bernoulli random variables (one per species).
This avoids known problems with the use of thresholds for determining whether or
not a species will be present at a site [[refs]]. We represented SDMs in this
analysis using random forests, a common approach in the species distribution
modeling literature based on fitting hundreds of independent regression trees to
randomly-perturbed versions of the data [[refs; would be good to provide a
source that says these are a good choice]]. Each of the 500 trees in our
species-level random forests used a different Monte Carlo estimate of the
observer effects as a predictor variable.

@@

Variance in species richness within sites was lower than predicted by several
common count models, such as the Poisson or binomial (i.e. richness was
underdispersed), so these count models would have had difficulty fitting the
data. Since richness had a relatively large mean ([[N]]) and all models already
produce continuous richness estimates (due to summing probabilities in the
(J)SDM case), we defined model performance for all models in terms of continuous
Gaussian errors instead of using discrete count distributions. When a model was
run multiple times for the purpose of correcting for observer effects, we used
the mean of those runs' point estimates as our final point estimate and we
calculated the uncertainty using the law of total variance (i.e.  the average of
the model runs' variance, plus the variance in the point estimates).

We evaluated the performance of each model for forecasting using the data for
each year between `r settings$last_train_year` and `r settings$end_yr`. We used
three metrics for evaluating performance: 1) root-mean-square error (RMSE) to
determine how far, on average, the models' predictions were from the observed
value; 2) the 95% prediction interval coverage to determine how well the models
predicted the range of possible outcomes; and 3) deviance (i.e. negative 2 times
the Gaussian log-likelihood) as an integrative measure of fit incorporating good
point estimates, precision, and coverage.  In addition to looking at forecast
performance in general we evaluated how performance changed as the time horizon
of forecasting increased by plotting performance metrics against the forecast
horizon.

@@

When comparing forecasts for richness across sites all methods performed well
(Figure 2; R^2 XX-XX%). However stacked and joint SDMs and the macroecological
model both failed to successfully forecast the highest richness sites resulting
in a notable clustering of predicted values near ~60 species and the poorest
performance in the cross-site comparison (R^2 XX-XX%).

For most combinations of metric and forecast time lag, the best-performing model
for forecasting within-site variability was either the "average" baseline model
or the ARIMA model [[Not sure what the bit about within-site variability means
here. Also, didn't Naive usually win when observer effects weren't corrected
for?]] (Figure 3). In general, each model's error increased more-or-less
linearly with time, although the slopes and initial error rates varied across
models (Figure 3). The site-level models (i.e. the baselines and the ARIMA
methods) tended to do very well (Figure 3) on short time horizons, especially in
terms of RMSE.

In general, the models produced confidence intervals that were too narrow,
indicating overconfident predictions. Of these, the random forest-based SDM
stack was the most overconfident, with observed richness values frequently
falling well outside their confidence limits (Figure 3C). As noted elsewhere
[@harris2015], this overconfidence is a product of the assumption in stacked
SDMs that errors in the species-level predictions are independent. The JSDM and
macroecological model has the best coverage, though each model still
underestimated the possible variation in richness. The "naive" model
was the only model whose confidence intervals were too wide (Figure 3), which
can be attributed to the rapid rate at which these intervals expand as the time
lag increases (Figure 1).

RMSE and deviance produced qualitatively similar results in most cases, with the
notable exception of the stacked SDMs (Figure 3A and 3B). Deviance, unlike RMSE,
depends on the full predictive distribution of a model, including the model's
estimated uncertainty. As a result, the stacked SDMs' deviance was notably
higher than the next-worst model even though its RMSE was not unusual. On the
other hand, the GBM-based "macroecological" model and especially the mistnet
JSDM had reasonably well-calibrated uncertainty estimates, even in the presence
of uncorrected observer biases (Figure 3C). This prevented their deviance from
rising as quickly as the site-level models as the observer population evolved
and resulted in better comparative performance for these models in terms of
deviance (Figure 3B).

Accounting for the uncertainty associated with differences among observers
resulted in improvements to all metrics of fit (Figure 4). The one exception was
RMSE of the "Naive" model, which remained unchanged. The improvements in point
estimates primarily resulted from large improvements in a small number of
forecasts, presumably ones where a large difference between observers resulted
in a large shift in predicted richness values that the models could not
otherwise accommodate. Most forecasts exhibited only small changes in RMSE.
However these changes were sufficient to change the rank ordering of model
performance. For example, the "naive" model generally had the lowest average
error when observer effects were not explicitly accounted for, switching to the
"average" and ARIMA model once observer differences were corrected. Improvements
in coverage were notable for all models except Naive and JSDM, meaning that
incorporating observer effects resulted in improved quantification of
uncertainty. The lack of change in the mistnet JSDM seems to indicate that its
latent variables were already accounting for observer-related uncertainty. These
differences meant that, in terms of deviance, neither the JSDM nor the
macroecological model model benefitted as much from the as the site-level models
did from the observer corrections. As a result, the average and ARIMA models
maintained a small average advantage over the JSDM in terms of deviance at
longer time scales, rather than falling behind after about five years.
